---
title: Connecting R to PostgreSQL on Linux
author: Mark Nielsen
date: '2018-07-07'
slug: connecting-r-to-postgresql-on-linux
categories:
  - Programming
tags:
  - RJDBC
  - odbc
  - RPostgreSQL
---

Connecting to databases is a critical piece of data anlaysis in R. In most analytic roles the data we consume is going to be found in databases. Of these some of the most common are SQL databases like MS SQL Server, PostgreSQL, and Oracle in addition to many others. In this how-to blog, I'll walk you through the major steps of configuring your machine and R to be able to connect to a PostgreSQL Server database from R on Ubuntu.  Similar steps can be followed to set up connections to other databases, however, driver installation and configuration will likely be slightly different.

### RPostgreSQL Setup
The first step in setting up a connection to a PostgreSQL database is to first download the PostgreSQL header files and static library, `libpq-dev`.  In order to do this on Ubuntu open the terminal and install it using the following command:
```{bash eval=FALSE}
sudo apt-get install libpq-dev
```

Once the `libpq-dev` package is installed the next step is to install the `RPostgreSQL` package in R. If you need to authenticate, I highly recommend the `getPass` package which will prompt you for your password. RStudio also has a `.rs.askForPassword()` function that works similar to the `getPass()` function, but it relies on using RStudio.  I've confirmed that `getPass` works in bash, emacs, and RStudio. So however you submit your R code it will work the same. 
```{r RPostgreSQL, eval=FALSE}
# Install the package in R
install.packages("RPostgreSQL")
```

```{r}
library(RPostgreSQL)
library(getPass)

pgdrv <- dbDriver(drvName = "PostgreSQL")

db <-DBI::dbConnect(pgdrv,
                    dbname="postgres",
                    host="localhost", port=5432,
                    user = 'postgres',
                    password = getPass("Enter Password:"))

# Write to database
DBI::dbWriteTable(db, "mtcars", mtcars)

DBI::dbDisconnect(db)
```
Perfect!  Your database connection should be working simply by adding the proper arguements in your `dbConnect()` function. You may need to tweak the host, port and user based on your PostgreSQL server setup.

### Install the `odbc` Package in R
In case you are a fan of `odbc`, the next section will walk you through the steps of creating your database connection via `odbc`.

In the past I have used the `RODBC` package but recently I have found that the `odbc` package plays much nicer with other database tools like `DBI` and `dbplyr`.  Plus it has very similar syntax to the `RJDBC` package and for consistency sake I've made the switch.

Once again, the first step is to install the necessary debian packages.  In this case we need to install the `unixodbc` and `unixodbc-dev` packages and the `odbc-postgresql` driver.
```{bash eval=FALSE}
# Install the unixODBC library
apt-get install unixodbc unixodbc-dev

# PostgreSQL ODBC Drivers
apt-get install odbc-postgresql
```


#### Set up connection with connection string
```{r odbc}
db <- DBI::dbConnect(odbc::odbc(),
                     Driver = "PostgreSQL Unicode",
                     Database = "postgres",
                     UserName = "postgres",
                     Password = getPass("Enter Password:"),
                     Servername = "localhost",
                     Port = 5432)
```

#### Set up connection with DSN
1. Make sure the `/etc/odbcinst.ini` has the drivers set up.  This should have been configured automatically when installing `odbc-postgresql` with `apt-get`.  This is what it would look like:
```{}
[PostgreSQL Unicode]
Driver = psqlodbca.so
Setup = libodbcpsqlS.so
Debug = 0
CommLog = 1
UsageCount = 1
```

2. Now define your DSN by modifying the odbc.ini file:
```{}
[PostgreSQL]
Driver = PostgreSQL Unicode
Database = postgres
Servername = localhost
UserName = postgres
Password = postgres
```

3. Connect to your database by referencing your DSN name specified in the square brackets of the `odbc.ini` file:
```{r odbc-dsn}
# Connect to the database
db <- dbConnect(odbc::odbc(), "PostgreSQL")

# Pull the Data into an R dataframe
DBI::dbGetQuery(db,"SELECT * FROM MTCARS")

# Close the Connection
DBI::dbDisconnect(db)
```

### Set up connection with RJDBC
```{r RJDBC}
install.packages('RJDBC')

library(RJDBC)

db <- DBI::dbConnect(RJDBC::JDBC("org.postgresql.Driver","postgresql-42.2.2.jar"),
               url = "jdbc:postgresql://localhost:5432/postgres",
               user = "postgres",
               password = getPass("Enter Password:"))


# Pull the Data into an R dataframe
DBI::dbGetQuery(db,"SELECT * FROM MTCARS")

# Close the Connection
DBI::dbDisconnect(db)

```


