---
title: Connecting R to PostgreSQL on Linux
author: Mark Nielsen
date: '2018-07-07'
slug: connecting-r-to-postgresql-on-linux
categories:
  - Programming
tags:
  - RJDBC
  - odbc
  - RPostgreSQL
---

Connecting to databases is a critical piece of data anlaysis in R. In most analytic roles the data we consume is going to be found in databases. Of these some of the most common are SQL databases like MS SQL Server, PostgreSQL, and Oracle in addition to many others. In this how-to blog, I'll walk you through the major steps of configuring your machine and R to be able to connect to a PostgreSQL Server database from R on Ubuntu using the `RPostgreSQL`, `odbc`, and `RJDBC` packages in R.  Similar steps can be followed to set up connections to other databases, however, driver installation and configuration will likely be slightly different.

### 1 - RPostgreSQL Package Setup
The first step in setting up a connection to a PostgreSQL database is to first download the PostgreSQL header files and static library, `libpq-dev`.  In order to do this on Ubuntu open the terminal and install it using the following command:
```{bash eval=FALSE}
sudo apt-get install libpq-dev
```

Once the `libpq-dev` package is installed the next step is to install the `RPostgreSQL` package in R. If you need to authenticate, I highly recommend the `getPass` package which will prompt you for your password. RStudio also has a `.rs.askForPassword()` function that works similar to the `getPass()` function, but it relies on using RStudio.  I've confirmed that `getPass` works in bash, emacs, RStudio, and when knitting your Rmd files. So however you submit your R code it will work the same. 
```{r RPostgreSQL, eval=FALSE}
# Install the package in R
install.packages("RPostgreSQL")
```

```{r}
library(RPostgreSQL)
library(getPass)
pgdrv <- dbDriver(drvName = "PostgreSQL")

db <-DBI::dbConnect(pgdrv,
                    dbname="postgres",
                    host="localhost", port=5432,
                    user = 'postgres',
                    password = getPass("Enter Password:"))

# Write to database
DBI::dbWriteTable(db, "mtcars", mtcars)

DBI::dbDisconnect(db)
```
Perfect!  Your database connection should be working simply by adding the proper arguments in your `dbConnect()` function. You may need to tweak the host, port and user based on your PostgreSQL server setup.

### 2 - odbc Package Setup
In case you are a fan of `odbc`, the next section will walk you through the steps of creating your database connection via `odbc`.

In the past I have used the `RODBC` package but recently I have found that the `odbc` package plays much nicer with other database tools like `DBI` and `dbplyr`.  Plus it has very similar syntax to the `RJDBC` package and for consistency sake I've made the switch.

Once again, the first step is to install the necessary debian packages.  In this case we need to install the `unixodbc` and `unixodbc-dev` packages and the `odbc-postgresql` driver.
```{bash eval=FALSE}
# Install the unixODBC library
apt-get install unixodbc unixodbc-dev

# PostgreSQL ODBC Drivers
apt-get install odbc-postgresql
```

#### Set up connection with connection string
Okay we are now ready to connect via odbc.  Note the slight difference in the names of the arguments of the `dbConnect()` function.
```{r odbc}
db <- DBI::dbConnect(odbc::odbc(),
                     Driver = "PostgreSQL Unicode",
                     Database = "postgres",
                     UserName = "postgres",
                     Password = getPass("Enter Password:"),
                     Servername = "localhost",
                     Port = 5432)
```

#### Set up connection with DSN
If you don't want to have to worry about defining each of these arguments each time you connect to PostgreSQL via odbc, you can define the configuration in your `odbcinst.ini` file.  The following steps walk you through the process:

1. Make sure the `/etc/odbcinst.ini` has the drivers set up.  This should have been configured automatically when installing `odbc-postgresql` with `apt-get`.  This is what it would look like:
```{}
[PostgreSQL Unicode]
Driver = psqlodbca.so
Setup = libodbcpsqlS.so
Debug = 0
CommLog = 1
UsageCount = 1
```

2. Now define your DSN by modifying the odbc.ini file:
```{}
[PostgreSQL]
Driver = PostgreSQL Unicode
Database = postgres
Servername = localhost
UserName = postgres
Password = postgres
```

3. Connect to your database by referencing your DSN name specified in the square brackets of the `odbc.ini` file:
```{r odbc-dsn}
# Connect to the database
db <- dbConnect(odbc::odbc(), "PostgreSQL")

# Pull the Data into an R dataframe
DBI::dbGetQuery(db,"SELECT * FROM MTCARS")

# Close the Connection
DBI::dbDisconnect(db)
```
Now you are ready to begin your analysis with your data!


### 3 - RJDBC Package Setup
Finally, the last way to configure a connection to the PostgreSQL database can be done via the `RJDBC` package.  The first step in this configuration is to download the jdbc jar file from [here](https://jdbc.postgresql.org/download.html). I've put this in my home directory, `~`, and will reference this file in the `JDBC()` function below.  Once you have the jar file you can install the `RJDBC` package in R.
```{r RJDBC, eval=FALSE}
install.packages('RJDBC')
```

Now you are ready to connect. Once again, notice the slight tweaks to the arguments of the `dbConnect()` function.  Because I'm defining the `url` argument with the host, port and database name, there is no need for these additional arguments.
```{r}
library(RJDBC)

db <- DBI::dbConnect(RJDBC::JDBC("org.postgresql.Driver","~/postgresql-42.2.2.jar"),
               url = "jdbc:postgresql://localhost:5432/postgres",
               user = "postgres",
               password = getPass("Enter Password:"))


# Pull the Data into an R dataframe
DBI::dbGetQuery(db,"SELECT * FROM MTCARS")

# Close the Connection
DBI::dbDisconnect(db)

```

Alright! We've walked through several different configurations in connecting to a PostgreSQL database on Ubuntu.  You'll only need one of these setups, but I think it's nice to understand each of your options so you can create the best setup that works for you and/or your organization.